Building an SSI Cluter with GNU/Hurd

### 2019-05-31 libpager/paper.tex
close to finishing 'Building an SSI Cluster' paper

\usepackage{nopageno}
\usepackage{vmargin}
\usepackage{comment}
\usepackage{xspace}
\usepackage[hyperfootnotes=false]{hyperref}
\usepackage{mdframed}
\usepackage{xcolor}

@@

% Shortcuts just for this document
\def\libpager{{\tt libpager}\xspace}
\def\netmsg{{\tt netmsg}\xspace}
\def\pagersync{{\tt pager_sync()}\xspace}
\def\pagerwritepage{{\tt pager_write_page()}\xspace}
\def\pagerreadpage{{\tt pager_read_page()}\xspace}
\def\moinit{{\tt memory_object_init}\xspace}
\def\moready{{\tt memory_object_ready}\xspace}
\def\morequest{{\tt memory_object_data_request}\xspace}
\def\mosupply{{\tt memory_object_data_supply}\xspace}
\def\modataerror{{\tt memory_object_data_error}\xspace}

@@

Mach's memory management was specifically designed to support
multi-node clusters.  Mach's model of memory management is that memory
is managed by user-space processes, and system memory is a cache of
that memory.  Any user-space process that implements Mach's memory
management protocol can be mapped into a process's address space.
When the process attempts to access memory, a page fault is triggered
which causes Mach to request the page from the user-space process
using a well documented protocol.  The faulting thread remains
suspended until the page has been supplied to the kernel.

@@

On one side, \libpager implements the Mach memory management protocol
interacting with the Mach kernel.  On the other side, \libpager
presents an API to the server program, which is responsible for
producing the pages, as follows:

@@

@@

@@

discard dirty pages!

 flush and return dirty (and precious) pages.

\item {\tt pager_get_error()}

@@

The current kernel implementation seems to request pages one at
a time, as the process attempts to access them.

If we're servicing a page read-only to multiple clients, where do we
get the data from?  If our local kernel has a copy, we almost
certainly want to get it there, since it's just a manner of flipping
bits in a hardware page table to obtain a new "copy" of the data.
Otherwise, we probably want to get the data from the local disk, but
if the disk is very busy and we've got fast network connections, we
might want to get that data from another node in the cluster.

Mach's current memory object API makes this a lot more difficult.  You
can't just request that the kernel hand you back an unmodified page,
except by flagging it "precious".  That's not too bad if the kernel
and the memory manager are both local on one node, but if you do this
over the network, you're sending unnecessary copies of the memory
pages around.

How do we figure out which client is the local kernel?  One possible
solution is for \netmsg to listen on /servers/netmsg for an RPC that
accepts a send right and identifies which node currently hosts the
corresponding receive right.  \cite{interfaces} defined the {\tt
  norma_port_location_hint} RPC that might be useful.

[[ the kernel uses different control ports for different memory objects ]]

  If the kernel uses the same control port for all of its pager
  operations, then we wouldn't have to make this query for every memory
  object; a single ext2fs translator, for example, could globally
  identify the location of all of the control ports, one for each node,
  and this information could be shared amongst all of that translator's
  pager objects.

DISTRIBUTED LIBPAGER.  We could use a shim process on remote nodes, to
allow requests for an existing page go to a node that holds a copy of
that page, instead of to the node with the disk.  Two ways I can
imagine this.  Either ext2fs forks off multiple processes on different
nodes, or a mechanism is developed to detect when a port is remote,
figure out which node it (currently) resides on, and then \libpager can
fork a process and push it to the other node, invisible to ext2fs
proper.

@@

  Currently, we read it again with {\tt pager_read_page()}, but if the
  local kernel has a copy, it's probably faster to ask the kernel for
  a copy.  We don't have a good way of detecting which client is the
  local kernel, and the Mach memory management API doesn't provide a
  way to request a copy of an unmodified, non-``precious'' page
  from the kernel.

@@

The primary data structures for mediating contention between clients
are the ACCESSLIST and the WAITLIST in the pagemap.

Putting a client on an empty WAITLIST triggers lock requests to all
clients on the ACCESSLIST, with the exception of the requesting client
itself, if it's already got read access and is requesting write
access.  Once the ACCESSLIST drops to zero (or one, in the exceptional
case just described), then the original request can be satisfied.

In the meantime, additional clients can be added to the WAITLIST,
which basically triggers nothing.  The rule is that if the WAITLIST is
not empty, then we're already trying to revoke access to the clients
on ACCESSLIST.  Remember, there's currently no guarantee that a client
will have access for any period of time at all.

When a client with write access returns a page, we look to see if
anything is on the WAITLIST.  If so, the first client(s) on WAITLIST
get the pages.  If not, we flag that a write is in progress and start
it.  When it completes, we check the WAITLIST again to see if anything
is now waiting.  If so, we service it.  If not, we discard the data
and notify the user that the page has been evicted.

If multiple clients with READ access attempt to unlock near
simultaneously, will the data_unlock logic will trigger multiple lock
requests (a set for each client attempting to unlock)?

initial client goes on WAITLIST only when:
1. client has requested any access and page is PAGINGOUT, or
2. client has requested any access and another client has WRITE access
3. client has requested WRITE access (data request or unlock) and another client has any access
4. client has requested any access and is waiting for the page to page in
     READ access requested - only READ clients on ACCESSLIST
     WRITE access requested - ACCESSLIST empty

@@

Mach kernels are allowed to proactively flush pages unless the
``precious'' flag is specified, so we can't be certain that a client
on a page's ACCESSLIST actually possesses that page.  All we know for
sure is that a client not on ACCESSLIST does not possess the page.

Examining the Mach kernel code for m_o_lock_request, we see that the kernel might block
while looping over the pages (and unlocks the memory object while doing so), so for a
multi-page lock request, we need to consider the possibility that part of the page range
might be processed, then other kernel operations happen (including VM page operations)
before the remaining part of the page range is processed and the lock request completes.

@@

\begin{itemize}
\item five clients have read access to page 50 and are on its ACCESSLIST (WAITLIST is empty)
\item pager_flush is called on pages 0-100
\item page 50 flushes on client A
\item client A re-requests access to page 50, and it is supplied (ACCESSLIST and WAITLIST don't change)
\item the flushes complete
\item only client A now has access to page 50 and should be alone on its ACCESSLIST
\end{itemize}

@@

PRECIOUS PAGES

Flagging a page precious (so the kernel always returns it to us) isn't that big a deal
for a local client, because ``returning the page'' just means flipping bits in the VM tables,
but it's a lot heavier operation for a remote client, since the page will be transmitted
across the network unnecessarily.

Flagging all of our pages ``precious'' is tempting.  It provides us with a mechanism
for retrieving a local, in-memory page that we want to serve out over the network
without having to read it again from disk.  It also simplifies the lock completed
logic, since we don't have to worry about whether the page got returned or not
when we get the lock completed message -- precious pages always get returned.

@@

client_request_read_access(client, start_page, end_page)
  ASSERT: page absent in client
  sends m_o_data_request
client_request_write_access(client, start_page, end_page)
  ASSERT: page absent in client
  sends m_o_data_request
client_request_unlock(client, start_page, end_page)
  ASSERT: page present in client with read access
  sends m_o_data_unlock
client_flush(client, start_page, end_page)
  ASSERT: pages present
  if pages have WRITE access, modify data and data_return
  if pages have READ access and are precious, data_return
client_service_message(client)
  read message
  data_supply - save pointers, read/write and precious (reply if requested)
  data_lock - send messages, update pointers, reply if requested
  data_error - ??

Another possible bug scenario (``bug3''): a single client holds two
pages with write access.  It returns one (perhaps because \pagersync
was called) which triggers a call to \pagerwritepage.  While
\pagerwritepage is running, both pages are returned in a single
m_o_data_return message.  Since the first page is still writing, we
sleep to wait for it to finish.  Then the second page is requested
again.  We {\it should} serve the copy that's queued to write, but if
we're not careful we'll call \pagerreadpage on the second page.

Test for BUG 3: (one client, two pages A and B)
  AB2 T4 (the read completes) TA3 A5 (the sync is processed) TAB2 A5 (the sync is processed)
    B2 (page request) T4 (the read completes)
    (now we need to close things out and see that data got discarded)

data should always be modified if possible (whenever a client's got WRITE access)

It was an interesting idea, but didn't quite pan out as well as I had
hoped.  For one thing, the operations can't be generated with equal
probability.  Client operation 5 (service a message) has to occur
fairly frequently, or a queue of pending messages will build up.
Likewise, translator operation 4 (complete a pending operation), also
has to occur fairly frequently.  On the other hand, client operation 5
can't run if there are no pending messages, and translator operation 4
can't run if there are no pending requests.

@@

, but can also be implemented using the {\tt
  tmpfs} translator (a ramdisk), which is linked with \libpager.

unionfs can possibly be used to
overlay the various /run/shm's on each other.

@@

The linux scheduler (sched-domains.txt) occasionally balances
load between scheduling domains.

\begin{comment}
\section{Blocking Behavior of {\tt mach_msg()}}

An issue with Mach's design caused me no little bit of grief.
Mach's {\tt mach_msg()} (as I read the documentation),
allows for non-blocking sends: according to \cite{interfaces}:

\begin{quote}
MACH_SEND_TIMEOUT

The {\it timeout} argument should specify a maximum time (in
milliseconds) for the call to block before giving up. If the message
can’t be queued before the timeout interval elapses, then the call
returns MACH_SEND_TIMED_OUT. A zero timeout is legitimate.
\end{quote}

In fact, a call to {\tt mach_msg()} can block indefinitely, even if a
{\it timeout} value of zero is specified!  The problem occurs when the
message is an RPC directed at the kernel which in turns triggers other
RPCs.  The memory manager is a big offender here.  Sending a {\tt
  vm_map} message to the kernel will cause the kernel to send a
\moinit message to the memory manager, then block waiting for a reply.
If both the {\tt vm_map} and the \moinit were handled by a single
thread in \netmsg, then that thread would block indefinitely,
requiring considerable additional complexity in \netmsg to ensure that
additional threads are always available in case calls to {\tt
  mach_msg()} block.

The consensus on the Hurd mailing list leaned in favor of this being
legitimate behavior, reading the documentation as guaranteeing that
the timeout applied to the send, and not to the underlying RPC.  In
any event, changing this behavior would require quite a bit of work on
the Mach kernel.

\end{comment}

@@

Both \netmsg and the new \libpager have been successfully
tested in a virtual machine running Hurd, but additional work is
required before Hurd can be used as a viable SSI cluster operating
system.

@@

@@

\item m_o_data_return could specify what permissions the kernel still holds on the page

@@

\item m_o_lock_completed message could specify what kind of lock was completed

@@

@@

SHUTDOWN CODE

Old \libpager would flush everything (no special flags set), then
destroy the receive right and ignore any final messages.  This created
a minor race condition where clients could request a page while the
flush was processing, which would then get dropped.

>>>>

\usepackage{nopageno}
\usepackage{vmargin}
\usepackage{comment}
\usepackage{xspace}
\usepackage[hyperfootnotes=false]{hyperref}
\usepackage{mdframed}
\usepackage{xcolor}
\usepackage{enumitem}
\usepackage{amssymb}

@@

% Shortcuts just for this document
\def\libpager{{\tt libpager}\xspace}
\def\netmsg{{\tt netmsg}\xspace}
\def\pagersync{{\tt pager_sync()}\xspace}
\def\pagerflush{{\tt pager_flush()}\xspace}
\def\pagerwritepage{{\tt pager_write_page()}\xspace}
\def\pagerunlockpage{{\tt pager_unlock_page()}\xspace}
\def\pagerreadpage{{\tt pager_read_page()}\xspace}
\def\moinit{{\tt memory_object_init}\xspace}
\def\moready{{\tt memory_object_ready}\xspace}
\def\morequest{{\tt memory_object_data_request}\xspace}
\def\moreturn{{\tt memory_object_data_return}\xspace}
\def\molockrequest{{\tt memory_object_lock_request}\xspace}
\def\molockcompleted{{\tt memory_object_lock_completed}\xspace}
\def\mosupply{{\tt memory_object_data_supply}\xspace}
\def\modataerror{{\tt memory_object_data_error}\xspace}

@@

Mach's memory management was specifically designed to support
multi-node clusters.  Mach's model of memory management is that memory
is managed by user-space processes, and system memory is a cache of
that memory.  Any user-space process that implements Mach's memory
management protocol can be mapped into a process's address space.
When the process attempts to access memory, a page fault is triggered
which causes Mach to request the page from the user-space process
using a well documented protocol.  The faulting thread remains
suspended until the page has been supplied to the kernel.
The current kernel implementation requests pages one at
a time, as the process attempts to access them, with no attempt
to anticipate future page accesses.

@@

\libpager acts as a translator between the Mach kernel and a
server program that parses the on-disk structure of a file system.
On one side, \libpager implements the Mach memory management protocol
interacting with the Mach kernel.  On the other side, \libpager
presents an API to the server program, which is responsible for
producing the pages.  The APIs primary functions are as follows:

@@

This routine is called when a page that was previously retrieved
for read-only access must now be promoted to read-write access.

@@

This routine is similar to {\tt pager_sync()}, but it requests that
the kernel(s) discard their copies of the pages after returning them.
Nothing prevents the kernels from immediately re-requesting copies
of the pages, but those new requests will trigger calls to
\pagerreadpage.

@@

This routine instructs the kernel to discard dirty pages without
bothering to save them!  Thus, the ``sync'', ``return'', and ``flush''
routines cover all possible combinations of two binary options:
``discard pages'' and ``return pages'':

\begin{tabular}{lcc}
& discard & return \\
{\tt pager_sync()} & & \checkmark \\
{\tt pager_return()} & \checkmark & \checkmark \\
{\tt pager_flush()} & \checkmark &
\end{tabular}

@@

@@

  Currently, we read it again with {\tt pager_read_page()}, but if the
  local kernel has a copy, it's probably faster to ask the kernel for
  a copy, since it's just a manner of flipping bits in a hardware page
  table to obtain a new ``copy'' of the data.  Otherwise, we probably
  want to get the data from the local disk, but if the disk is very
  busy and we've got fast network connections, we might want to get
  that data from another node in the cluster.

  We don't have a good way of detecting which client is the local
  kernel, and the Mach memory management API doesn't provide a way to
  request a copy of an unmodified, non-``precious'' page from the
  kernel.  You can't just request that the kernel hand you back an
  unmodified page, except by flagging it ``precious''.  That's not too
  bad if the kernel and the memory manager are both local on one node,
  but if you do this over the network, you're sending unnecessary
  copies of the memory pages around.

  How do we figure out which client is the local kernel?  One possible
  solution is for \netmsg to listen on {\tt /servers/netmsg} for an
  RPC that accepts a send right and identifies which node currently
  hosts the corresponding receive right.  \cite{interfaces} defined
  the {\tt norma_port_location_hint} RPC that might be useful.

  If the kernel used the same control port for all of its pager
  operations (it currently doesn't), then we wouldn't have to make
  this query for every memory object; a single ext2fs translator, for
  example, could globally identify the location of all of the control
  ports, one for each node, and this information could be shared
  amongst all of that translator's pager objects.

@@

The primary data structures for mediating contention between clients
are the ACCESSLIST and the WAITLIST in the pagemap.  The ACCESSLIST
can have multiple clients on it, if they all have read-only access.
Write access can only be issued to one client at a time.

A client requesting access goes on a previously empty WAITLIST only when:
\begin{enumerate}[noitemsep, itemsep=2pt, topsep=0pt]
\item The client has requested any access and we have to start a page in.  This happens when
  \begin{itemize}[label={}]
  \item READ access requested - only READ clients on ACCESSLIST
  \item WRITE access requested - ACCESSLIST empty
  \end{itemize}
\item The client has requested any access and we're waiting for the page to page out
\item The client has requested any access and another client (on ACCESSLIST) has WRITE access
\item The client has requested WRITE access and other clients (on ACCESSLIST) have any access
\end{enumerate}

Putting a client on an empty WAITLIST triggers lock requests to all
clients on the ACCESSLIST, with the exception of the requesting client
itself, if it's already got read access and is requesting write access
(an unlock request).  Once the number of clients on the ACCESSLIST
drops to zero (or one, in the unlock case), then the original request
can be satisfied.

In the meantime, additional clients can be added to the WAITLIST,
which basically triggers nothing.  The rule is that if the WAITLIST is
not empty, then we're already trying to revoke access to the clients
on ACCESSLIST.  Thus, the wheels have already been set in motion for
the WAITLIST to be serviced, and we are free to add additional clients
to the end of the WAITLIST.  Remember, there's currently no guarantee
that a client will have access for any period of time at all, so
there's no time period when any clients will be on WAITLIST without an
attempt being made to revoke existing access.

When a client with write access returns a page, we look to see if
anything is on the WAITLIST.  If so, the first client(s) on WAITLIST
get the pages.  If not, we flag that a write is in progress and start
it by calling \pagerwritepage.  When it completes, we check the
WAITLIST again to see if anything is now waiting.  If so, we service
the request.  If not, we discard the data and notify the server (if
it requested notifications) that the page has been evicted.

@@

Mach kernels are allowed to proactively flush pages unless the
``precious'' flag is specified, so we can't be certain that a client
on a page's ACCESSLIST actually possesses that page.  All we know for
sure is that a client not on ACCESSLIST does not possess the page.  This
makes it somewhat difficult to compute ACCESSLIST after a \pagerflush.

Examining the Mach kernel code for \molockrequest, we see that the kernel might block
while looping over the pages (and unlocks the memory object while doing so), so for a
multi-page lock request, we need to consider the possibility that part of the page range
might be processed, then other kernel operations happen (including VM page operations)
before the remaining part of the page range is processed and the lock request completes.

@@

\begin{itemize}
\item five clients have read access to page 50 and are on its ACCESSLIST (WAITLIST is empty)
\item \pagerflush is called on pages 0-100
\item page 50 flushes on client A
\item client A re-requests access to page 50, and it is supplied (ACCESSLIST and WAITLIST don't change)
\item the flushes complete
\item only client A now has access to page 50 and should be alone on its ACCESSLIST
\end{itemize}

@@

Flagging all of our pages ``precious'' (so the kernel always returns
it to us) is tempting.  It provides us with a mechanism for retrieving
a local, in-memory page that we want to serve out over the network
without having to read it again from disk.  It also simplifies the
flushing logic above, since we don't have to worry about whether the
page got returned or not when we get the lock completed message --
precious pages always get returned.  Flagging a page ``precious''
isn't that big a deal for a local client, because ``returning the
page'' just means flipping bits in the VM tables, but it's a lot
heavier operation for a remote client, since the page will be
transmitted across the network unnecessarily, so we don't do this.

@@

Another possible bug scenario (``bug3''): a single client holds two
pages with write access.  It returns one (perhaps because \pagersync
was called) which triggers a call to \pagerwritepage.  While
\pagerwritepage is running, both pages are returned in a single
\moreturn message.  Since the first page is still writing, we
sleep to wait for it to finish.  Then the second page is requested
again.  We {\it should} serve the copy that's queued to write, but if
we're not careful we'll call \pagerreadpage on the second page.

The test code would test for this bug as follows: (one client, two pages A and B)

AB2 T4 (the read completes) TA3 A5 (the sync is processed) TAB2 A5 (the sync is processed)
B2 (page request) T4 (the read completes)

It was an interesting idea, but didn't quite pan out as well as I had
hoped.  For one thing, the operations can't be generated with equal
probability.  Client operation 5 (service a message) has to occur
fairly frequently, or a queue of pending messages will build up.
Likewise, translator operation 4 (complete a pending operation), also
has to occur fairly frequently.  On the other hand, client operation 5
can't run if there are no pending messages, and translator operation 4
can't run if there are no pending requests, so there is some
synchronization that has to occur.  Finally, I'm not sure that
\libpager can be fully tested using anything other than actual Mach
kernels.

@@

This implementation services all shared memory requests by directing
them to a single node.  Obviously, this is not ideal.  The {\tt
  unionfs}\footnote{{\tt unionfs} is not part of the stock Hurd
  distribution.  Its source is available from
  \href{https://www.gnu.org/software/hurd/hurd/translator/unionfs.html}{\tt
    https://www.gnu.org/software/hurd/hurd/translator/unionfs.html}}
translator allows directories from different nodes to be overlaid on
top of each other, and could possibly be used to overlay the various
{\tt /dev/shm}'s on each other.  Then, a new shared memory segment
would be created on the original process's node, but would immediately
become visible to all other nodes.  A more sophisticated translator
could be developed that would dynamically balance shared memory
management among the nodes.

@@

The same mechanisms used for cross-node forks could also be used to
permit process migration.  After using cross-node memory maps to
create a duplicate process on another node (without any threads
running), the threads of the original process would be stopped, the
Mach port rights transferred using \netmsg, duplicate threads would be
created and started.

Some investigation would be required to see if Mach can cleanly
transfer thread state like this.  For example, I'm not sure how Mach
would save state for a thread currently blocked in a {\tt mach_msg()}
system call.  I think that it saves the user-space state with the
instruction pointer set to the system call instruction.  Restarting
from this state would rerun the system call, which would create
duplicate messages if, for example, the system call was the {\tt
  mach_msg_send()} variant.

@@

Both \netmsg and the new \libpager have been successfully
tested in a virtual machine running Hurd, but their operation
is unreliable and additional work is
required before Hurd can be used as a viable SSI cluster operating
system.

@@

\item Other translator improvements

\item \libpager integration with {\tt tmpfs}

{\tt tmpfs}, the Hurd ramdisk, currently throws {\tt SIGABRT} if any
\libpager functions are called.
Since {\tt tmpfs} is the most obvious choice to mount on
{\tt /dev/shm}, fixing this is a high-priority, low-difficulty task.

Once we have the ability to run processes using distributed
shared memory and distributed scheduling, we should enhance our file
system translators to support replicated copies of files.  This
suggests the need for a...

\item distributed \libpager

We could use a shim process on remote nodes, to allow requests for an
existing page go to a node that holds a copy of that page, instead of
to the node with the disk.  Two ways I can imagine this.  Either
ext2fs forks off multiple processes on different nodes, or a mechanism
is developed to detect when a port is remote, figure out which node it
(currently) resides on, and then \libpager can fork a process and push
it to the other node, invisible to ext2fs proper.

@@

\item \moreturn could specify what permissions the kernel still holds on the page

@@

\item \molockcompleted message could specify what kind of lock was completed

@@

\small

@@
